Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-05 manuscript/manuscript.Rmd

@@

>>>>

“In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states at some time in the future.  It is important to distinguish this objective from the extensive previous literature on “early warning signs” of critical transitions [eg as reviewed in @Scheffer2015] which has sought to answer only a categorical question: is the system approaching a critical transition? More recent work by @Bury2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not). These are important results with considerable success, but do not answer the questions of when a shift might occur or how large that transition will be. In light of increasing calls for the importance of such quantitative, probabilistic forecasts [@Deitz2018], we focus on evaluating the potential and limits to prediction in critical transition scenarios.”

@@

**Selecting timescales**
In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Figure 1.  
For the Hopf bifurcation, forecast begins at t=100 and extends to t =200, for the saddle node, forecast begins at t=250 and extends to t = 500, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t = 250.  
While much attention is often paid to the number of data points in training or testing data, it is essential to realize that these are only meaningful relative to the specific process in question.  
Thus, in each case we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario.  
For instance, if the stochastic scenario was restricted to a much shorter timescale used in the hopf case, few replicate simulations would experience a transition at all.  
If it were made much longer, most of the timeseries would be spent post-transition. 
Likewise, if the forecast for the Hopf scenario was extended much further into the future under the current parameterization, the system experiences a homoclinic bifurcation at which the population collapses to 0.  
Using different length timescales allows us to consider the three different forecasting tasks illustrated in Figure 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
These three critical transitions are fundamentally different processes, there is no perfect apples-to-apples parameterization for each that allows the transition to unfold in a way that gives precisely the same time windows.

time at which to start the forecast in each scenario relative to the timing of the bifurcation event (in the case of Hopf and Saddle Node), and relative to the expected distribution of transitions in the stochastic case. The rate at which the dynamical process unfolds in each case depends on the parameter values of the model in question.  At much
